---
output: html_document
editor_options: 
  chunk_output_type: console
---
\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

# Load your packages
library(here)
library(tidyverse)
library(lubridate)

# Set your ggplot theme

tb_el_theme <- theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(color = "darkblue"),
    panel.background = element_rect(fill = "gray"),
    legend.position = "top")

theme_set(tb_el_theme)

# Load your datasets

soil_hp_2013_raw <- read.csv(here("./Data/Raw/Soil_Hydraulic_Properties/SHPs2013.csv"))
soil_hp_2015_raw <- read.csv(here("./Data/Raw/Soil_Hydraulic_Properties/SHPs2015.csv"))
soil_hp_2017_raw <- read.csv(here("./Data/Raw/Soil_Hydraulic_Properties/SHPs2017.csv"))

soil_ps_2013_raw <- read.csv(here("./Data/Raw/Soil_Particle_Size/psaBF2013.csv"))
soil_ps_2015_raw <- read.csv(here("./Data/Raw/Soil_Particle_Size/psaBF2015.csv"))
soil_ps_2017_raw <- read.csv(here("./Data/Raw/Soil_Particle_Size/psaBF2017.csv"))

soil_bd_2013_raw <- read.csv(here("./Data/Raw/Soil_Bulk_Density/bulkdensity2013_N10.csv"))
soil_bd_2015_raw <- read.csv(here("./Data/Raw/Soil_Bulk_Density/bulkdensity2015.csv"))
soil_bd_2017_raw <- read.csv(here("./Data/Raw/Soil_Bulk_Density/bulkdensity2017.csv"))
```


# Rationale and Research Questions

Wildfires are becoming more severe, and climate change is creating warmer and
drier conditions in many places around the globe, making them more wildfire prone.
Depending on their severity, wildfire can have significant and negative impacts
on physical properties of soil, which directly relates to how much water an
area can hold. 

The 2013 Black Forest Fire in Colorado's El Paso County covered 14,280 acres
and destroyed 489 homes. It is one of the most destructive forest fires in the
state's history. Following the fire, researchers collected soil data at various
sites with varying burn severity. 

We seek to better understand how the wildfire and its different intensities
impacted the soil's physical characteristics. 

\newpage

# Dataset Information

Data Source: Soil-physical and soil-hydraulic properties as a function of burn severity for 2013, 2015, and 2017 in the area affected by the 2013 Black Forest Fire, Colorado USA (ver. 2.0, June 2021)

Retrieved from: https://www.sciencebase.gov/catalog/item/5b8452f5e4b05f6e321b629b

Variables used: Soil dry bulk density, soil particle size, soil-hydraulic properties

Data range: 2013-07-23 to 2017-06-19


\newpage

# Exploratory Analysis 

```{r}

```


\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
